A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility

### 2019-06-18 paper.Rmd

For a piece of computational research to be minimally reproducible, it requires three distinct elements:

1. Code  
2. Computing environment  
3. Data

The first two of these challenges have largely been solved (REFERENCE).

@@

- **Who** collected it
- **What** is the data
- **When** was it collected
- **Where** was it collected
- **How** is was collected

@@

> A high-level summary, ideally combined with text explaining the structure and nature of the data set, also helps to explain an unfamiliar data set to researchers who want to reproduce analyses or reuse the data.

@@

**Variable names** are short, descriptive names with no spaces or special characters. For example, "job_position", "faculty_level", and "years_at_company". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, 0 = no, 1 = yes, and 0 = male and 1 = female. These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 =yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the codebook, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown" or -99 [@White2013; @Broman2017].

@@

Data dictionary tables should be placed in the README and presented as a table. Every data dictionary should also be provided in its raw form in the repository, for example, saved as a CSV (see Figure XX.).

@@

Data with a license clearly establishes rules on how everyone can modify, use, and share data. Without a license, these rules are unclear, and can lead to problems with attribution and citation.

It can be overwhelming to try and the right license for a use case. Two licenses that are well suited for data sharing are:

@@

The data dictionary provides _human readable_ information on the data. To ensure data types are preserved - dates are dates, names are characters, and so on - there needs to be some form of _machine readable_ metadata. An excellent standard for metadata is Table Schema written by frictionless data [@fritionless-table-schema]. Say for a dataset called "demographics" with three variables (Table \@ref(tab:table-machine-readable)) and JSON equivalent in Figure \@ref(fig:json-code).

@@

This contains fields such as path, and a schema with subfields name and type, for each variable. It also provides information for, licensing and features such as line breaks, and delimiters. It is built on JSON (JavaScript Object Notation), a lightweight, human-friendly, machine readable data-interchange format. Table schema is baked into formats such as csvy [@csvy], an extended `csv` format, which has additional front matter in a YAML format using Table Schema.

@@

The data used in the data analysis should be provided in a folder called `data`.
Ideally, the data should be in "Tidy Data" format [@Wickham2014], where tidy data contains variables in columns, and observations in rows. Contrasting `raw data`, `tidy data`/`analysis data` should be in an easily readable plain-text format, such as CSV, tab separated, or semicolon separated. Binary or proprietary formats are discouraged in favor of interoperability.

@@

Tooling for producing these this information speeds up the process of sharing data. Machine-readable metadata that can be indexed by google is created using the `dataspice` package [@dataspice]. To help create codebooks, we recommend the `codebook` package in R [@Arslan2019], which also generates machine readable metadata. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with DataPackageR [@DataPackageR], which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to [Frictionlessdata's tabular data package spec [@frictionlessdata-data-pkg].

@@

To demonstrate the variety of datasets and documentation practices, we now discuss three datasets of varying quality.

@@

One example of a very laborious, long-term study is the 50-hectare plot on Barro Colorado Island in Panama. Researchers at the field station have censused every single tree in a pre-defined plot 7 times over the past 30 years. More than a million trees have been counted. The data and metadata however and hard to reach. To obtain the data, one must fill out a form and agree to terms and conditions which require approval and authorship prior to publication. The biggest challenge with this study is that the data are stored on a personal FTP server of one of the authors. While the data are available in CSV and binary Rdata formats, the data storage servers do not have any metadata, README files or codebooks. A separate, public archive hosts the metadata (https://repository.si.edu/handle/10088/20925) in a PDF file (https://repository.si.edu/bitstream/handle/10088/20925/RoutputFull.pdf?sequence=1&isAllowed=y) that describe the fields.

@@

Datasets obtained from sensors such as meteorological data are typically easy to prepare and understand, since sensors measure specific features. This means the description of data type happens upstream at the instrument-design level, and flows down to data collection. The telescope data from the Sloan Digital Sky Survey (SDSS) is a great example of this.

It is not possible to summarise the contents of the data, but briefly, the SDSS data includes photometric and spectroscopic information obtained from a 2.5m telescope at Apache Point Observatory in New Mexico, operating since 2000,s producing over 200Gb of data every day [@sdss-one-paper; @sdss-four-paper; @sdss-website]. This data has been used to produce the most detailed, three dimensional maps of the universe ever made.

@@

```{r sdss-raw-data-screenshot, fig.cap = "Screenshot of Raw data avaialbe through DR15 FITS."}
knitr::include_graphics(here::here("figures",
                                   "sdss-raw-data-screenshot.png"))
```

@@

- Metadata in the machine readable standard, SCHEMA
- Data in multiple formats (e.g., .csv and .fits)
- Previously released data available in archives
- Data is available in a database
- Entire datasets available for download
- Raw data before processed into database is also available

As an example, the optical and spectral data can be accessed in FITS data format, or even as a plain CSV, the first few rows shown in Table \@ref(tab:spectra-tab). An example plot in Figure \@ref(fig:sdss-spectra)A shows the Wavelength against Flux as produced on the SDSS website, while Figure \@ref(fig:sdss-spectra)B. The fact that figure Figure \@ref(fig:sdss-spectra)A is reproduced in Figure \@ref(fig:sdss-spectra)B demonstrates great reproducibility.

@@

```{r sdss-spectra, fig.cap = "Spectra image of wavelength against flux from SDSS from two sources. Part A shows the example image taken from SDSS website. Part B shows the replicated figure using data provided as CSV, demonstrating great reproducibility."}

@@

Metadata from the data cannot be contained in the CSV file, but is available in the URL the csv was downloaded from:

```
http://dr15.sdss.org/optical/spectrum/view/data/format=csv?
plateid=1678&mjd=53433&fiberid=425&reduction2d=v5_7_0
```

@@

* **plateid** is the plate that they use to explore a given section of night sky. It uniquely identifies a "plug plate" used to eollcet the data
* **mjd** is an integer that marks the "modified julian date" of the night when observations was carried out.
* **fiberid** is an integer that marks the fiber number used (the fiber optic cable used to transmit light to the sensor).
* The **version** marked as `v5_7_0` is the pipeline version used for BOSS in DR12.

@@

The SDSS provides data at different levels of complexity for different degrees of understanding. It is a huge project involving many people, and a staggering effort to prepare. Despite this, it is still very easy to understand. This is a further reflection on the idea that high effort  can create highly understandable data, (see Figure \@ref(fig:fig-effort-understanding)). The impact of this data is high, having changed the way we understand the past, present and future of the universe itself. This impact is surely due to the care and thought put into making the data accessible and public. It is a statement of what is possible with good data sharing practices.

@@

1. **Decide whether publishing is appropriate for your data** 
It is critical to publish your data if they are the basis for a peer-reviewed research publication. To be broadly useful, your data must be deposited in a permanent archive (to ensure that it does not disappear from a ephemeral location such as your university website), along with useful metadata so that someone outside the project team can understand what the data mean. However, this level of effort is not always critical for all data science efforts. For small-scale projects such as ones where one might generate simulated datasets, this level of curation is highly unnecessary. Here making data available in a transient location like GitHub or a website is sufficient.

2. **Include a README** file with your data archive. README files have a long history in software (https://medium.com/@NSomar/readme-md-history-and-components-a365aff07f10) and are named so that the ASCII systems capital letters filename would show this file first, making it the obvious place to put relevant information. In the context of datasets, READMEs are particularly useful when there are no reliable standards. The best practice here is to have one readme per dataset, with names that easily associate with the corresponding data files. In addition to metadata about the data such as variable names, units of measurement and code books, the readme should also contain information on licenses, authors/copyright, a title, and dates and locations where the data were collected. Additionally keywords, author identifiers such as ORCID, and funder information would be useful.

@@

4. **Provide a machine readable format** for your data. When possible, provide machine readable metadata that map on to the open standards of [schema.org](https://schema.org/) and [JSON-LD](https://en.wikipedia.org/wiki/JSON-LD). These metadata provide all of the information described in the readme best practices (rule 2), but in a machine readable way and include much of the same information such as name, description, spatial/temporal coverage etc. One way to test if your metadata are machine readable is to use Google's structured testing data (https://search.google.com/structured-data/testing-tool/u/0/) to verify the integrity of the metadata.

@@

6. **Provide [open source?] scripts** used to clean data from rawest form into tidy/analysis ready format. raw data is often unusable without further processing (data munging or data cleaning), which are the steps necessary to detect and clean inaccurate, incomplete records, and missing values from a dataset. While datasets can be cleaned interactively, this approach is often very difficult to reproduce. It is a better practice to use scripts to batch process data cleaning. Scripts can be verified, version controlled and rerun without much overhead when mistakes are uncovered.  These scripts describe the steps taken to prepare the data, which helps explain and document any decisions taken during the cleaning phase. These should ideally operate on some raw data stored in the "data-raw" folder (rule 5).

@@

9. **Only add a citation if your data has a DOI**. A citation only makes sense when your data has a DataCite compliant DOI, which is automatically provided when data is published in repositories like Zenodo and Dryad. Even then a citation may not accrue references, but without a DOI, it is guaranteed not to.

10. **Stick with simple data formats** to ensure long-term usefulness of datasets in a way that is not tied to transient technologies (such as ever changing spreadsheet standards), store the data as plain text where possible. This can take the form of comma/tab separated files in most cases. This will ensure transparency and future proof your data.

@@

The open science literature has been advocating the importance of data sharing for a long time. However many of these articles appeal to the broader benefits of data sharing but rarely talk about the practical considerations of making data available alongside code. Trivial reproducibility has always relied up on the idea that as long as code exists on a platform such as GitHub and clearly lists open source dependencies, one should be able to run the code at a future time. Containerization efforts such as Docker have made it even easier to capture system dependencies and rapidly launch containers. But true reproducibility requires not just code but also data.

Over the years researchers have made data available in a multitude of unreliable ways, including offering downloads from university websites, FTP repositories, and password protected servers, all of which are prone to bit rot. In rare cases where the data are deposited in a permanent archive, insufficient metadata and missing data processing scripts have made it harder and time intensive to map these to code.  Our goal here is to describe a variety of ways in which a researcher can ship data, analogous to how easy it has become to push code to services like GitHub. The choice of technology and method all depends on the nature of the data, size, and complexity. Once the issue of access is solved, we also discuss how much metadata to include in order to make the data useful. Ultimately there is a tradeoff to consider on a case by case basis. Long-term studies with complex protocols require structured metadata which more standardized datasets such as those coming from sensors require far less effort.

@@

# Morgue

## Challenges - Data format

Data that is easy to use should avoid being in a proprietary format that requires special software to read.  For example, a more basic format such as CSV (Comma Separated Value) should be preferred over proprietary data formats, such as SPSS or SAS custom data types, or binary formats, as it is more reliable, although it can sometimes be slower to read in and out, and harder to share due to size.

## Challenges -  Understanding tradeoffs

To improve data impact it needs more time to prepare so it is easier to understand.

Another challenge is in the gross total amount of documentation. Too much documentation can be overwhelming; too little and you don't know anything. For example, compare a folder with dozens of data files and several papers, to a single CSV file. The first is too much, and understanding it requires a high time cost. The second is too little; we don't know anything about what the data is or why it was collected. The challenge is to provide the right level of detail, somewhere between these two extremes. This changes depending on the impact of the data, and the task at hand.

There are many features to consider when documenting, sharing, and publishing data. What format to store the data in? How do you describe the variables, observations, and collection methods? Do you include raw data? What about data used in analysis? What about metadata for the variables? And then there's licenses, and long term storage. It can be hard to know where to start, but together they all help make the data easier to understand. These features are not clearly described in the scientific literature in the context of including them alongside research.

The research created from the 50 hectare plot, discussed in section \@ref(ex-data-1) - this dataset was challenging to understand without special domain expertise.

Ultimately, creating good documentation has the same challenge as good writing: it takes time, and it can be hard to know when you are done.

It can be hard to find the right amount of time to spend. Ideally, you want the gains in understanding not to be outweighed by the time taken to improve them.

Impact is a difficult quantity to measure.

To help explain these descriptive qualities of the data sharing, we discuss three datasets and their relationship of **Effort to prepare**, and **Ease of understanding** in section 5.

One of the easiest solutions to making data available alongside an analysis is to include it in the same version controlled repository as the analysis scripts. This works well with the data file sizes are in the order of bytes to megabytes. When a reader clones the repository, the scripts and notebooks will be able to read all data files from relative paths with no further modifications to the code itself. Assuming the data would be useful beyond the context of the analysis, it would be beneficial to also archive it in a long-term repository.

>>>>

For a piece of computational research to be minimally reproducible, it requires three distinct elements: 1) Code;  2) Computing environment, and 3) Data. The first two of these challenges have largely been solved (REFERENCE).

@@

- **Who** collected it
- **What** is the data
- **When** was it collected
- **Where** was it collected
- **Why** it was collected
- **How** is was collected

@@

@@

**Variable names** are short, descriptive names with no spaces or special characters. For example, "job_position", "faculty_level", and "years_at_company". **Variable labels** are longer descriptions of variables. For example "University Job Position", "University Faculty Position Level", and  "Number of Years Worked at University" [@McGill-codebook; @Arslan2019]. **Variable codes** apply to categorical (factor) variables, and are the values for their contents. For example, `0 = no, 1 = yes` for one variable, and `0 = inside` and `1 = outside`. These should be consistent across similar variables to avoid problems where `0 = yes` for one variable, and `1 = yes` in another. Date variables should have consistent formatting. For example, all date information could be in format "YYYY-MM-DD", and this should not be mixed with "YYYY-DD-MM". **Missing data** are values that should have been observed, but were not. The code for missingness should be documented in the codebook, and should nominally be `NA`. If the reason for missingness is known it should be recorded. For example, censored data, patient drop out, or measurement error can have different values, such as "unknown" or -99 [@White2013; @Broman2017].

@@

Data dictionary tables should be placed in the README and presented as a table. Every data dictionary should also be provided in its raw form in the repository, for example, saved as a CSV.

@@

Data with a license clearly establishes rules on how everyone can modify, use, and share data. Without a license, these rules are unclear, and can lead to problems with attribution and citation. It can be overwhelming to try and the right license for a use case. Two licenses that are well suited for data sharing are:

@@

The README, and data dictionary provides _human readable_ information on the data. To ensure data types are preserved - dates are dates, names are characters, and so on - there needs to be some form of _machine readable_ metadata. An excellent standard for metadata is Table Schema written by frictionless data [@fritionless-table-schema]. For example, a dataset called "demographics" with three variables is shown in Table \@ref(tab:table-machine-readable), which has the Java Script Object Notation (JSON) equivalent in Figure \@ref(fig:json-code).

@@

This contains fields such as path, and a schema with subfields name and type, for each variable. It also provides information for licensing and features such as line breaks, and delimiters. It is built on JSON (JavaScript Object Notation), a lightweight, human-friendly, machine readable data-interchange format. Table schema is baked into formats such as csvy [@csvy], an extended `csv` format, which has additional front matter in a YAML format using Table Schema.

@@

The data used in the data analysis should be provided in a folder called `data`. Ideally, the data should be in "Tidy Data" format [@Wickham2014], where tidy data contains variables in columns, and observations in rows. Contrasting "raw data", "analysis ready data" should be in an easily readable plain-text format, such as CSV, tab separated, or semicolon separated. Binary or proprietary formats are discouraged in favor of interoperability, as it requires special software to read, even if it is sometimes slower to read in and out, and harder to share due to size.

@@

Tooling for producing data documentation information speeds up the process of sharing data. Machine-readable metadata that can be indexed by google is created using the `dataspice` package [@dataspice]. To help create codebooks, we recommend the `codebook` package in R [@Arslan2019], which also generates machine readable metadata. Codebooks are implemented in other software such as STATA, which provides a "codebook" command. Data can be packaged up in a "data package" with DataPackageR [@DataPackageR], which provides tools to wrap up data into an R package, while providing helpers with MD5sum checks that track versioning. Note that is different to Frictionlessdata's tabular data package spec [@frictionlessdata-data-pkg].

@@

We now explore the variety of documentation practices of a few selected datasets.

@@

One example of a very laborious, long-term study is the 50-hectare plot on Barro Colorado Island in Panama. Researchers at the field station have censused every single tree in a pre-defined plot 7 times over the past 30 years. More than a million trees have been counted. The data and metadata however are hard to reach. To obtain the data, one must fill out a form and agree to terms and conditions which require approval and authorship prior to publication. The biggest challenge with this study is that the data are stored on a personal FTP server of one of the authors. While the data are available in CSV and binary Rdata formats, the data storage servers do not have any metadata, README files or codebooks. A separate, public archive hosts the metadata (https://repository.si.edu/handle/10088/20925) in a PDF file (https://repository.si.edu/bitstream/handle/10088/20925/RoutputFull.pdf?sequence=1&isAllowed=y) that describe the fields.

@@

Datasets obtained from sensors such as meteorological data are typically easy to prepare and understand. This is because sensors measure specific features, so the description of data type happens upstream at the instrument-design level, and flows down to data collection. The telescope data from the Sloan Digital Sky Survey (SDSS) is a great example of sensor data.

The SDSS data includes photometric and spectroscopic information obtained from a 2.5m telescope at Apache Point Observatory in New Mexico, operating since 2000, producing over 200Gb of data every day [@sdss-one-paper; @sdss-four-paper; @sdss-website]. This data has been used to produce the most detailed, three dimensional maps of the universe ever made.

@@

```{r sdss-raw-data-screenshot, fig.cap = "Screenshot of Raw data avaialbe through DR15 FITS.", out.width = "60%", fig.align = "center"}
knitr::include_graphics(here::here("figures",
                                   "sdss-raw-data-screenshot.png"))
```

@@

- Metadata in the machine readable standard, SCHEMA
- Data in multiple formats (e.g., .csv and .fits, database)
- Previously released data available in archives
- Entire datasets available for download
- Raw data before processed into database is also available

For example, the optical and spectral data can be accessed in FITS data format, or even as a plain CSV, the first few rows shown in Table \@ref(tab:spectra-tab). Wavelength plotted against Flux is shown in Figure \@ref(fig:sdss-spectra), with \@ref(fig:sdss-spectra)A showing the output from the SDSS website, and Figure \@ref(fig:sdss-spectra)B showing the output from the CSV. The fact that figure Figure \@ref(fig:sdss-spectra)A is virtually replicated in Figure \@ref(fig:sdss-spectra)B demonstrates great reproducibility.

@@

```{r sdss-spectra, fig.cap = "Spectra image of wavelength against flux from SDSS from two sources. Part A shows the example image taken from SDSS website. Part B shows the replicated figure using data provided as CSV, demonstrating great reproducibility.", out.width = "90%", fig.align = "center"}

@@

Metadata from the data cannot be contained in the CSV file, but is available in the URL the csv was downloaded from: `http://dr15.sdss.org/optical/spectrum/view/data/format=csv?`
`plateid=1678&mjd=53433&fiberid=425&reduction2d=v5_7_0`

@@

* **plateid:** the unique plate used to explore a given section of night sky. 
* **mjd:** an integer that marks the "modified julian date" of the night of observation.
* **fiberid:** an integer of the fiber optic cable used to transmit light to the sensor.
* The **version** marked as `v5_7_0` is the pipeline version used for BOSS in DR12.

@@

The SDSS provides data at different levels of complexity for different degrees of understanding. It is a huge project involving many people, and a staggering effort to prepare. Despite this, it is still very easy to understand. This is a further reflection on the idea that high effort  can create highly understandable data, (see Figure \@ref(fig:effort-understanding)). The impact of this data is high, having changed the way we understand the past, present and future of the universe itself. This impact is surely due to the care and thought put into making the data accessible and public. It is a statement of what is possible with good data sharing practices.

@@

1. **Decide whether publishing is appropriate for your data** 
It is critical to publish your data if they are the basis for a peer-reviewed research publication. To be broadly useful, your data must be deposited in a permanent archive, to ensure that it does not disappear from a ephemeral location such as your university website. It shoudl also contain useful metadata so that someone outside the project team can understand what the data mean. However, this level of effort is not always critical for all data science efforts. For small-scale projects such as ones where one might generate simulated datasets, this level of curation is highly unnecessary. Here making data available in a transient location like GitHub or a website is sufficient.

2. **Include a README** file with your data archive. README files have a long history in software (https://medium.com/@NSomar/readme-md-history-and-components-a365aff07f10) and are named so that the ASCII systems capital letters filename would show this file first, making it the obvious place to put relevant information. In the context of datasets, READMEs are particularly useful when there are no reliable standards. The best practice here is to have one README per dataset, with names that easily associate with the corresponding data files. In addition to metadata about the data such as variable names, units of measurement and code books, the README should also contain information on licenses, authors/copyright, a title, and dates and locations where the data were collected. Additionally keywords, author identifiers such as ORCID, and funder information would be useful.

@@

4. **Provide a machine readable format** for your data. When possible, provide machine readable metadata that map on to the open standards of [schema.org](https://schema.org/) and [JSON-LD](https://en.wikipedia.org/wiki/JSON-LD). These metadata provide all of the information described in the README best practices (rule 2), but in a machine readable way and include much of the same information such as name, description, spatial/temporal coverage etc. One way to test if your metadata are machine readable is to use Google's structured testing data (https://search.google.com/structured-data/testing-tool/u/0/) to verify the integrity of the metadata.

@@

6. **Provide [open source?] scripts** used to clean data from rawest form into tidy/analysis ready format. Raw data is often unusable without further processing (data munging or data cleaning), which are the steps necessary to detect and clean inaccurate, incomplete records, and missing values from a dataset. While datasets can be cleaned interactively, this approach is often very difficult to reproduce. It is a better practice to use scripts to batch process data cleaning. Scripts can be verified, version controlled and rerun without much overhead when mistakes are uncovered.  These scripts describe the steps taken to prepare the data, which helps explain and document any decisions taken during the cleaning phase. These should ideally operate on some raw data stored in the "data-raw" folder (rule 5).

@@

9. **Only add a citation if your data has a DOI**. A citation only makes sense when your data has a DataCite compliant DOI, which is automatically provided when data is published in repositories like Zenodo and Dryad. Although a citation may not accrue references, without a DOI it is guaranteed not to.

10. **Stick with simple data formats** to ensure long-term usefulness of datasets in a way that is not tied to transient technologies (such as ever changing spreadsheet standards), store the data as plain text (e.g., CSV) where possible. This can take the form of comma/tab separated files in most cases. This will ensure transparency and future proof your data.

@@

The open science literature has been advocating the importance of data sharing for a long time. Many of these articles appeal to the broader benefits of data sharing, but rarely talk about the practical considerations of making data available alongside code. Trivial reproducibility has always relied upon the idea that as long as code exists on a platform such as GitHub, and clearly lists open source dependencies, one should be able to run the code at a future time. Containerization efforts such as Docker have made it even easier to capture system dependencies, and rapidly launch containers. But true reproducibility requires not just code, but also data.

Over the years researchers have made data available in a multitude of unreliable ways, including offering downloads from university websites, FTP repositories, and password protected servers, all of which are prone to bit rot. In rare cases where the data are deposited in a permanent archive, insufficient metadata and missing data processing scripts have made it harder and time intensive to map these to code.  Our goal here is to describe a variety of ways in which a researcher can ship data, analogous to how easy it has become to push code to services like GitHub. The choice of technology and method all depends on the nature of the data, and its size and complexity. Once the issue of access is solved, we also discuss how much metadata to include in order to make the data useful. Ultimately there is a tradeoff to consider on a case by case basis. Long-term studies with complex protocols require structured metadata, while more standardized datasets such as those coming from sensors require far less effort.

@@
